Testable normative theories

### 2020-04-18 simple.tex
minor
Reworked for one simple model overall

\section{Motivation}
TODO The model should be: (m, s) leadsto (t, sure) where m can be for example: after having presented s0, s1, s2, present s3, and tell the individual to wait for one week.

In decision theory and economy in general, preferences are generally considered as “intuitive”, as originating from an “immediate sensation”, following the terms used by \citet{von_neumann_theory_2004} (vNM) \citep{fishburn_retrospective_1989, cailloux_reasons_2017}. Though appropriate when studying the everyday behavior of individuals, this conception is insufficient to study the positions that individuals adopt towards difficult decision problems after careful thinking. In this article I define the concept of deliberated preferences to fill this gap: the deliberated preferences of an individual about a given decision problem represent the position of that individual towards that problem that is stable facing counter-arguments.

@@

\section{General normative theories}
A general normative theory is a tuple $(\alltopic, \allargs, I, (\ileadsto, i \in I), \gargs, \gleadsto)$.

@@

Then the normative criterion is: $t$ in DJ of $i$ when $\exists s \suchthat \forall s': (s', s) \ileadstost^i t$ (the theory provides a part of this guarantee under falsification, claiming for all $i$, the model makes the rest under validation for this $i$).

@@

The relation $\ileadstost$ must satisfy the two following properties, or axioms. (A1) $\ileadstost \subseteq \ileadstoe$
\footnote{Equivalently: $\ileadstosst \subseteq \ileadstose$, $\ileadstopst \subseteq \ileadstope$, $(\ar, \ar') \ileadstose ¬\prop ∨ (\ar, \ar') \ileadstope \prop$, $\ileadstopst \prop ⇒ ¬ (\ileadstosst ¬\prop)$, $\ileadstosst ¬\prop ⇒ ¬ (\ileadstopst \prop)$.} 
and (A2) $\forall \prop \in \alltopic: (\ar, \ar') \ileadstost (\prop, \text{sure}) ⇒ (\ar, \ar') \ileadstost (\prop, \text{possible})$
\footnote{Equivalently: $\ileadstose \subseteq \ileadstope$.}. 
Observe that (A1) can equivalently be formulated as $\forall \prop \in \alltopic: [(\ar, \ar') \ileadstost (\prop, \text{sure}) ⇒ (\ar, \ar') \ileadstoe (\prop, \text{sure})]$, or equivalently, $\forall \prop \in \alltopic: [(\ar, \ar') \ileadstost (\prop, \text{possible}) ⇒ (\ar, \ar') \ileadstoe (\prop, \text{possible})]$, and (A2) is equivalent to $\forall \prop \in \alltopic: [(\ar, \ar') \ileadstoe (\prop, \text{sure}) ⇒ (\ar, \ar') \ileadstoe (\prop, \text{possible})]$.

@@

\subsection{Deliberated preference}
The following important definition is required in order to define the deliberated preference of $i$. An argument $\ar$ is decisive for $\phi$, in short, $\phi\text{-dec}$ iff $\forall \ar' \in \allargs: (\ar', \ar) \ileadstost \phi$. 
\footnote{Equivalently: $\ar \ \phidec ⇔ [\ileadstostinv(\phi)](\ar) = \allargs$, and $\exists \ar \ \phidec ⇔ [\ileadstostinv(\phi)]^{-1}(\allargs) ≠ \emptyset$.}
I also say that $\ar$ is $\propsuredec$, $\proppossdec$, or $¬\proppossdec$, when $\phi=(\prop, \text{sure})$, $\phi=(\prop, \text{possible})$, or $\phi=(¬\prop, \text{possible})$, respectively.
\footnote{$\propsuredec ⇔ \forall \ar': (\ar', \ar) \ileadstosst \prop ⇔ \forall \ar': ¬[(\ar', \ar) \ileadstope ¬\prop]$, $\proppossdec ⇔ \forall \ar': (\ar', \ar) \ileadstopst \prop$. Note that “both” $≠$ “we donno”: in case “both”, we know how to prove that both are reasonable options, in the sense that neither is certain. We have excluded some possibilities (and some other model could disagree with the “both” conclusion).}

@@

Define Finite defense-$\wibeatse$-$\wibeatse$-subsets as: $\wibeatseinv(\ar) \subseteq \wibeatse(\args) ⇒ \wibeatseinv(\ar) \subseteq \wibeatse(\args')$. Finite defense-$\mbeats$-$\wibeatse$-subsets $⇏$ Finite defense-$\wibeatse$-$\wibeatse$-subsets. Consider $\ar_2 \mbeats^\text{fail} \ar_1$ (to be continued…)

\bibliography{simple}

\end{document}

>>>>

\section{Motivation}
In decision theory and economy in general, preferences are generally considered as “intuitive”, as originating from an “immediate sensation”, following the terms used by \citet{von_neumann_theory_2004} (vNM) \citep{fishburn_retrospective_1989, cailloux_reasons_2017}. Though appropriate when studying the everyday behavior of individuals, this conception is insufficient to study the positions that individuals adopt towards difficult decision problems after careful thinking. In this article I define the concept of deliberated preferences to fill this gap: the deliberated preferences of an individual about a given decision problem represent the position of that individual towards that problem that is stable facing counter-arguments.

@@

\section{Normative theories}
Sketch: define normative models, using questions, and making claims about DJ. A problem is that models specific to a narrow situation will not offer enough means for falsification. Then define a way of saying “we never need this argument”: either the model can prove phi without s0, or if it can’t, then phi does not hold, because I can play s1 against it. BUT what about a model for only one situation? Then we need to accept that $q$ can’t be falsified, and accept that it defines the individual (or just assume without proof that it does not).

A normative theory is a tuple $(\allprops, ¬, \allargs, I, (\ileadstost[i])_{i \in I}, q, (\gleadsto[q(i)])_{q(i) \in q(I)}, \gbeats)$,
with $\gargs \subseteq \allargs$, ${\ileadstost[i]} \subseteq (\allargs × \allargs) × \Phi$, ${\gleadsto[q(i)]} \subseteq \allargs × \Phi$, ${\gbeats} \subseteq \allargs × \allargs$,
where $\Phi = \allprops × \set{\text{possible}, \text{sure}}$.

To lighten notations, $\ileadstost[i]$ will generally be denoted by $\ileadstost$, and $\gleadsto[q(i)]$ by $\gleadsto$, with the understanding that these symbols result of applying functions of, respectively, $i$ and $q(i)$. When the symbol $i$ is used in a formula without being bound, the binding $\forall i \in I$ is implied.

As usual, $(\ar', \ar) \nileadstost \phi$ means $¬[(\ar', \ar) \ileadstost \phi]$.

The symbol $¬$ is a negation operator on $\allprops$, such that $\forall \prop \in \allprops: ¬¬\prop = \prop$ (this article also uses this symbol to negate classical logic propositions and as a negation operator on $\Phi$, which should create no confusion). Define $¬$ on $\Phi$ as follows, $\forall \prop \in \allprops$: $¬(\prop, \text{possible}) = (¬\prop, \text{sure})$ and $¬(\prop, \text{sure}) = (¬\prop, \text{possible})$. Observe that $¬¬\phi = \phi$.

Abusing notation, given $\ar \in \allargs$ and $\phi \in \Phi$, I write $\ar \ileadstost \phi$ to mean $[\forall \ar' \in \allargs: (\ar', \ar) \ileadstost \phi]$.

The set $\iphi \subseteq \Phi$ is the set of interest. 
When $(\prop, \text{possible}) \in \iphi$, $\prop$ is considered as a possibility in the deliberated judgment of $i$ (it is not rejected), and when $(\prop, \text{sure}) \in \iPhi$, $\prop$ is considered as a certainty in the deliberated judgment of $i$. TODO we should have $(\prop, \text{sure}) \in \iPhi ⇒ (¬\prop, \text{possible}) \notin \iPhi$.
Part of the content of $\Phi$ is known as a result of a correct normative and empirical analysis.

To ease intuition, let $(\ar', \ar) \ileadstoe \phi$ be equivalent to $(\ar', \ar) \nileadstost ¬\phi$, $\forall \ar', \ar, \phi$.

Given $\phi \in {\gleadsto}(\allargs)$, an argument $\ar \in \allargs$ is \emph{in favor of $\phi$}, $\ar \in \gargs[\phi]$, iff $\ar \in [\gbeatsinv]^{2k}({\gleadstoinv}(\phi))$ for some $k \in \N$, where $R^\alpha$ denotes $\alpha$ times the composition of the relation $R$ with itself. (Throughout the article, $\N$ includes zero.) Define $\gargs = \cup_{\phi \in {\gleadstoinv}(\allargs)} \gargs[\phi]$.

Define ${\ibeatse} \subseteq \allargs × \gargs$ such that $\ar' \ibeatse \ar$ iff $\exists \phi \suchthat \ar \in \gargs[\phi] \text{ and } (\ar', \ar) \ileadstoe ¬\phi$. 
Given $\phi \in {\gleadsto}(\allargs)$, define ${\ibeatse[\phi]} \subseteq \allargs × \gargs[\phi]$ such that $\ar' \ibeatse[\phi] \ar$ iff $(\ar', \ar) \ileadstoe ¬\phi$. 
Define ${\ibeatsst} \subseteq \gargs × \allargs$ such that $\ar' \ibeatsst \ar$ iff $\forall \phi \suchthat \ar' \in \gargs[\phi]: (\ar', \ar) \ileadstost \phi$.
\footnote{When both $\ar_1, \ar_2 \in \gargs$, the relationship is as follows. $\ar_2 \ibeatse \ar_1 ⇒ \ar_2 \nibeatsst \ar_1$, the converse failing only if $\ar_2$ has some claims that $\ar_1$ does not have (thus $\exists \phi \suchthat \ar_2 \in \gargs[\phi] \land \ar_1 \notin \gargs[\phi]$): $\ar_2 \nibeatsst \ar_1$ exactly when $\ar_2$ fails to argue decisively for some of its claims, and $\ar_2 \nibeatse \ar_1$ exactly when $\ar_1$ argues decisively for all of its claims.}

A normative theory starts with an axiom stating its normative adequacy.
\begin{axiom}[Normative adequacy of $\allargs$ and $\ileadstost$]
	A general normative theory is \emph{normatively adequate} iff $\forall \phi \in \Phi$: 
	\begin{equation}
		[\exists \ar \in \allargs \suchthat \ar \ileadstost \phi] ⇒ 
		\begin{cases}
			\phi \in \iphi,\\
			\lnot\phi \notin \iphi.
		\end{cases}
	\end{equation}
\end{axiom}

This notation is justified only if it is impossible that both $\phi \in \iphi$ and $\phi \notin \iphi$ be implied. \commentOC{TODO clarify.} The following conditions permit to obtain this result.

\begin{condition}[Always implies sometimes]
	\label{cl:ais}
	${\ileadstost} \subseteq {\ileadstoe}$.
\end{condition}

\begin{condition}[Sure implies possible]
	\label{cl:sip}
	$\forall \ar', \ar \in \allargs, t \in \allprops: (\ar', \ar) \ileadstost (\prop, \text{sure}) ⇒ (\ar', \ar) \ileadstost (\prop, \text{possible})$.
\end{condition}

\begin{proposition}
	$\forall \phi \in \Phi$: 
	\begin{equation}
		[\exists \ar \in \allargs \suchthat \ar \ileadstost \phi] ⇒ [\nexists \ar \in \allargs \suchthat \ar \ileadstost ¬\phi],
	\end{equation}
	and $\forall \prop \in \allprops$:
	\begin{equation}
		[\exists \ar \in \allargs \suchthat \ar \ileadstost (\prop, \text{sure})] ⇒ [\nexists \ar \in \allargs \suchthat \ar \ileadstost (¬\prop, \text{sure})].
	\end{equation}
\end{proposition}
\begin{proof}
	The first result obtains because $\ar \ileadstost \phi$ means that given any $\ar'$, $(\ar, \ar') \ileadstost \phi$, hence $(\ar, \ar') \ileadstoe \phi$ (using \cref{cl:ais}), or equivalently, $(\ar, \ar') \nileadstost ¬\phi$, thus no $\ar' \in \allargs$ satisfies $\ar' \ileadstost ¬\phi$. For the second one, a similar reasoning yields that $(\ar, \ar') \nileadstost (¬\prop, \text{possible})$, and using the contrapositive of \cref{cl:sip}, $(\ar, \ar') \nileadstost (¬\prop, \text{sure})$, thus no $\ar' \in \allargs$ satisfies $\ar' \ileadstost (¬\prop, \text{sure})$.
\end{proof}

\begin{definition}[Validity]
	A theory is \emph{valid} iff ${\gleadsto}(\allargs) \subseteq \iphi$.
\end{definition}

The following condition mandates that if the arguments used by the theory are not decisive, it knows a reason.
\begin{definition}[Answerability]
	$\forall \phi \in \Phi, \ar_1 \in \allargs, \ar_2 \in \gargs[\phi] \cap {\gbeatsinv}(\ar_1): (\ar_2, \ar_1) \ileadstoe ¬\phi ⇒ [\exists \ar_3 \in {\gbeatsinv}(\ar_2) \suchthat (\ar_3, \ar_2) \ileadstoe \phi]$.
\end{definition}

\begin{definition}[Bounded width]
	$\forall \ar \in \gargs: \card{\ibeatseinv(\ar)} ≤ b_w$.
\end{definition}

\begin{definition}[Bounded length]
	$\forall n \in \N$, if $n > b_l$, there is no sequence of arguments $(\ar_k)_{0 ≤ k < n}$ such that $[\exists \phi \suchthat \forall (0 ≤ k < 2n): \ar_{2k} \in \gargs[\phi]]$ and $(\ar_{k + 1}, \ar_k) \ileadstoe \phi$ and $(\ar_{k + 1}, \ar_k) \ileadstoe ¬\phi$.
\end{definition}

\begin{definition}[Reinstatement]
	A theory satisfies \emph{reinstatement} iff $\forall \ar_1, \ar_3 \in \gargs[\phi], \ar_2 \in \allargs: [\ar_1 \ileadstoe \phi, (\ar_2, \ar_1) \ileadstoe ¬\phi, (\ar_3, \ar_2) \ileadstost \phi] ⇒ (\ar_1 + \ar_3, \ar_2) \ileadstost \phi$.
\end{definition}

\begin{definition}[Operational validity]
	A theory is \emph{operationally valid} iff $\forall \phi \in {\gleadsto}(\allargs), \ar_0 \in \gargs[\phi], \ar_1 \in \allargs: (\ar_1, \ar_0) \ileadstoe ¬\phi ⇒ [\ar_1 \gbeats \ar_0] \land [\exists \ar_2 \in {\gbeatsinv}(\ar_1) \suchthat (\ar_2, \ar_1) \ileadstoe \phi]$.
\end{definition}

\begin{theorem}
	If a theory is answerable, satisfies reinstatement, has bounded width and length, and is operationally valid, it is valid.
\end{theorem}
\begin{proof}
	Given $\phi \in {\gleadsto}(\allargs), \args_2, \args_0 \subseteq \gargs$, define $\args_2$ as \emph{$\phi$-defending} $\args_0$ iff $\forall \ar_1 \in \allargs: [\exists \ar_0 \in \args_0 \suchthat (\ar_1, \ar_0) \ileadstoe ¬\phi] ⇒ [\exists \ar_2 \in \args_2 \suchthat (\ar_2, \ar_1) \ileadstoe \phi]$. Define $\args_2$ as \emph{decisively $\phi$-defending} $\args_0$ iff $\forall \ar_1 \in \allargs: [\exists \ar_0 \in \args_0 \suchthat (\ar_1, \ar_0) \ileadstoe ¬\phi] ⇒ [\exists \ar_2 \in \args_2 \suchthat (\ar_2, \ar_1) \ileadstost \phi]$.

	Fix $\phi \in {\gleadsto}(\allargs)$. For all $k \in \N$, define $\gargs[k] = [\gbeatsinv]^{2k}({\gleadstoinv}(\phi))$.
	Note that $\ar \in \gargs[\phi] ⇔ \exists k \in \N \suchthat \ar \in \gargs[k]$.

	Lemma: $\gargs[k+2]$ $\phi$-defends $\gargs[k]$. To prove this, fix $\ar_1 \in \allargs, \ar_0 \in \gargs[k] \suchthat (\ar_1, \ar_0) \ileadstoe ¬\phi$. We need to exhibit $\ar_2 \in \gargs[k+2] \suchthat (\ar_2, \ar_1) \ileadstoe \phi$. Indeed, we know that $\ar_1 \ibeatse[\phi] \ar_0$, and we can use Operational validity as $\ar_0 \in \gargs[\phi]$, thus, $\ar_1 \gbeats \ar_0$ and $\exists \ar_2 \in {\gbeatsinv}(\ar_1) \suchthat (\ar_2, \ar_1) \ileadstoe \phi$.

	\commentOC{I suppose that, therefore, $\gargs$ $\phi$-defends itself!}

	Pick any even $b > b_l$. Lemma: $\gargs[b] = \emptyset$, and $\gargs[b]$ defends $\gargs[b-2]$, thus, $\gargs[b-2]$ has no attacker.

	Th: $\gargs[4]$ decisively defends $\gargs[2]$ (easy if $j = 4$).
	Then: if $\gargs[4]$ decisively defends $\gargs[2]$, then anything $\gargs[2]$ defends, it decisively defends.
\end{proof}

Model: for a given $i$. Defends $\prop$. $\exists \ar_1 \in \args_\eta \suchthat \ar_1 \ileadstoe \prop$. Then, $\forall \ar_2 \in \gargs \suchthat (\ar_2, \ar_1) \ileadstoe ¬\prop: \exists \ar_3 \in \args_\eta \suchthat (\ar_3, \ar_2) \ileadstoe \prop.$

@@

Then the normative criterion is: $t$ in DJ of $i$ when $\exists s \suchthat \forall s': (s', s) \ileadstost t$ (the theory provides a part of this guarantee under falsification, claiming for all $i$, the model makes the rest under validation for this $i$).

\commentOC{Say that a theory supposedly draws on general principles, which justifies that it talks about multiple $\phi$, even though at this level, there is no connection between them.}

@@

@@

\subsection{Deliberated preference}
The following important definition is required in order to define the deliberated preference of $i$. An argument $\ar$ is decisive for $\phi$, in short, $\phi\text{-dec}$ iff $\forall \ar' \in \allargs: (\ar', \ar) \ileadstost \phi$. 
\footnote{Equivalently: $\ar \ \phidec ⇔ [\ileadstostinv(\phi)](\ar) = \allargs$, and $\exists \ar \ \phidec ⇔ [\ileadstostinv(\phi)]^{-1}(\allargs) ≠ \emptyset$.}
I also say that $\ar$ is $\propsuredec$, $\proppossdec$, or $¬\proppossdec$, when $\phi=(\prop, \text{sure})$, $\phi=(\prop, \text{possible})$, or $\phi=(¬\prop, \text{possible})$, respectively.
\footnote{$\propsuredec ⇔ \forall \ar': (\ar', \ar) \ileadstosst \prop ⇔ \forall \ar': ¬[(\ar', \ar) \ileadstope ¬\prop]$, $\proppossdec ⇔ \forall \ar': (\ar', \ar) \ileadstopst \prop$. Note that “both” ≠ “we donno”: in case “both”, we know how to prove that both are reasonable options, in the sense that neither is certain. We have excluded some possibilities (and some other model could disagree with the “both” conclusion).}

@@

Define Finite defense-$\wibeatse$-$\wibeatse$-subsets as: $\wibeatseinv(\ar) \subseteq \wibeatse(\args) ⇒ \wibeatseinv(\ar) \subseteq \wibeatse(\args')$. Finite defense-$\mbeats$-$\wibeatse$-subsets $⇏$ Finite defense-$\wibeatse$-$\wibeatse$-subsets. Consider $\ar_2 \mbeats^\text{fail} \ar_1$ (to be continued…)
\end{document}
